Spatial considerations in the analysis of biological conservation data: space is special

### 2017-01-11 paper_v3_EI.Rmd
Add files via upload

---
title: 'Spatial considerations in the analysis of biological conservation data: space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Carol Ximena Garzon-Lopez^2^, Paul Harris^3^, Duccio Rocchini^4^"
output: 
  pdf_document: 
    keep_tex: yes
geometry: margin=3cm
---

@@

# Normal Abstract
Geographically Weighted Regression (GWR) as first proposed by Brunsdon et al (1996) is a commonly used approach in many local geographical and biogeographical analyses. One of the of criticisms of the use of GWR is that local models may be subject to local collinearity between variables, even when none is observed globally. As yet none of the published research describing applications of GWR has addressed this issue. This paper does so. It applies GWR to the model suggested by Coudun et al (2006) to predict the presence of *Acer campestre* and to account for any spatial non-stationarity, as observed in many statistical patterns and relationships. The analysis tests for presence of local collinearity amongst predictor variables and applies a locally compensated GWR model where it is needed. The results describe the spatial variation of coefficient estimates predicting *Acer campestre* and compensates for any local collinearity amongst variables using a local ridge term. The paper discusses the need for *geography goggles* in  analyses of spatial and in so doing it encourages explicitly spatial ways of thinking. Such considerations are increasingly relevant as more and more data have a spatial component in the form of location, for instance from GPS, and all data are collected some*where*. These suggest the need for more informed approaches for analysing space and location than provided by standard statistical models.

# GEB Abstract
**Aim**: The aim of this research was to test the model suggested by Coudun et al (2006) that predicts the presence of *Acer campestre*, to explore species distributions models using Geographically Weighted Regression (GWR) as first proposed by Brunsdon et al (1996), and to take account of local collinearity amongst predictor variables. Local collinearity may occur even when none is observed globally. None of the many papers describing applications of GWR have addressed this issue.\
**Location**: UK \
**Time period**: 1980 and 2010\
**Major taxa studied**: *Acer Campestre*\
**Methods**: Data on the presence of *Acer Campestre* were downloaded from GBIF, and linked to data describing  autumn rainfall, wildness quality and potential evapotranspiration (PET). These data were aggregated over OS 10km squares. The analysis compares the results of predictive models from applying a standard OLS regression, a standard GWR and a locally compensated ridge GWR. A GWR diagnostics procedures was used to test for the presence of local collinearity and where found a locally compensated GWR model was applied.\
**Results**: The results show that the presence of *Acer campestre* is significantly associated with potential evapotranspiration and Wilderness Quality Index. The GWR analyses describe the spatial variation of coefficient estimates predicting *Acer campestre* and these are contrasted with the locally compensated GWR.\
**Main conclusions**: The paper discusses the need for *geography goggles* in  analyses of spatial and in so doing it encourages explicitly spatial ways of thinking. Such considerations are increasingly relevant as more and more data have a spatial component in the form of location, for instance from GPS, and all data are collected some*where*. These suggest the need for more informed approaches for analysing space and location than provided by standard statistical models.\

@@

Ecological data increasingly have locational attributes thanks to near ubiquitous GPS in measurement and observation devices. This paper emphasises the importance of explicitly considering the spatial properties of data and argues that location cannot be treated as just another variable, precisely because of the prevalence of spatial autocorrelation or spatial non-stationarity in data, variables and statistical relationships. Space is special and requires *geography goggles* to be worn in applied biogeographical / ecological analyses and their associated methods and models. In order to demonstrate the value of spatially explicit methods, this paper tests the regression model suggested by Coudun et al (2006) but does this using a Geographically Weighted (GW) framework (Brunsdon et al, 1996), in this case a GW Regression (GWR). Coudun et al (2006) found the presence of *Acer campestre* in France to be significantly related to two factors: rainfall and evapotranspiration. Here Coudon’s analysis is extended to account for a particular form of spatial dependence in the data which occurs when changes in properties of nearby features are found to be correlated. This contradicts the underlying assumptions of independence and stationarity made in many statistical analyses and inferences. The result is spatial autocorrelation or spatial non-stationarity when the statistical pattern or statistical relationship observed in one location differs from that in another. To explore these, this paper applies a GWR analysis to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* (field maple) presence to test for the presence of local, non-stationary relationships. GWR is directly concerned with modelling spatial non-stationarity  and in doing so will often indirectly account for any spatial autocorrelation (e.g. Harris etc al. 2011)

Spatially explicit approaches such as GWR reflect Tobler's First Law of Geography (Tobler, 1970). They seek to quantify *local* patterns in relationships and processes rather than global 'whole map' statistics. Typically,  local models are constructed from subsets of the data, for example by selecting data under a moving window or kernel, and local methods such as GWR have been used in a number of biodiversity and ecological studies. Recent examples include Roll et al (2015) who examined the spatial variation tree height as a predictor of species richness for different taxa, Wang et al (2016) who used to GWR to develop local models of net primary production in Chinese forests and Keith et al (2013) who sought to predict local species richness using a GWR analysis of environmental variables. However, local methods may be subject to collinearity even when it is not observed globally (Brunsdon et al 2012) because of the particular characteristics of the local subset. Collinearity will affect model reliability, result in inferential bias through unstable parameter estimates, inflated standard errors and difficulties in separating variable effects, etc (Dormann et al 2013; Meloun et al. 2002). Collinearity is commonly tested for in analyses of a-spatial data and may even be used to predict missing data. Despite the many research papers describing applications of GWR, as yet there are no instances of research that has considered the potential effects of local collinearity. As well as extending the analysis of Coudun et al (2006) using GWR, this paper explicitly tests for and compensates for locally collinearity where found, with dramatic differences between the coefficient estimates and the mapped GWR outputs without and without local compensations. The paper proceeds with a review of approaches for handling spatial auto-correlation (Section 2), Section 3 describes data and study area along with the issues of collinearity and local collinearity, before describing the results of a GWR analysis, a GW local collinearity diagnostics analysis and a locally compensated ridge regression in Section 4. Section 5 discusses the wider implications of the research int he context of the increasing availability of spatial data and some conclusions are drawn in Section 6.

@@

Understanding species distribution and spatial dependencies is a key concept in ecological research (Rocchini et al., 2015), in particular when dealing with biodiversity monitoring for conservation practices (Honrado et al., 2016). Several papers have dealt with the importance of explicitly taking space into account when trying to model the distribution of a certain species, and, overall, its change in space and time (see Pearman et al., 2008 and references therein). The description of the distribution of biodiversity at different spatial and temporal scales has long been the focus of ecology and biogeography. Reliable descriptions of species distributions are fundamental for conservation and research purposes (Dormann, 2007) and for conservation, where a lack of accuracy hamper the potential to provide effective tools for the development of spatially and temporally informed biodiversity management strategies (Guisan et al., 2014, Porfirio et al., 2014). However, when modelling species distributions, a number of statistical problems must be solved before spatial models can be built. These include the uncertainty related to the sampling of the modelled species (Rocchini et al., 2011), correlation among predictors (Dormann et al., 2012), potential variability over space of the predictors being used (Rocchini et al., 2016) and   spatio-temporal dependencies of the predictors (Zuur and Ieno, 2016). Among these, spatial non-stationarity and scale dependence have the greatest impact on the observed ecological processes and their relative patterns, such as the distribution of biodiversity over space (Foody, 2004).

The importance of spatial considerations has long been recognised in a number of disciplines: Fischer (1935) in crop science; Kolmogorov (1941), Gandin (1965) in meteorology; Krige (1951) and  Matheron (1963) in mining; Matérn (1960) in forestry; theoretical developments by Moran (1950) and Yaglom (1955); Berry and Marble (1968) and Chorley and Haggett (1967) in geography; and Legendre and Legendre (1991) in ecology.

The major issue with modelling spatial data is the lack of independence of the spatial objects and the assumption of stationarity in the processes being modelled. These are particularly true with socio-economic data, but are also true of bio-geographical, environmental and ecological data. Both concerns relate to Tobler's dictum (the first Law of Geography) that: '*everything is related to everything else, but near things are more related than distant things*' (Tobler 1970). However, in spite of numerous methodological advances in addressing such spatial modelling problems, policy related research is still routinely informed by non-spatial modelling practices, which are heroic at best, and ill-advised at worst. The lineage, from the late 1970s onwards, of such methodological advances can be loosely traced through spatial statistics texts from Journel and Huigbrechts (1978) to Cressie (1993), to Chilès and Delfiner (1999) and to Cressie and Wilke (2011).

Focusing on regression models, the main drawback in assuming observations are independent is that any spatial dependencies turn up in the residuals, as they are not explicitly added to the model. For area-based data, ways to account for this include Besag's (1974) conditionally autoregressive model or Anselin's (1988) spatially autoregressive model.  In both cases, a criterion is needed for determining 'nearby' and is often supplied by the entries in a matrix of adjacencies. Wall (2004) notes that spatial covariance structures are implied by such matrices, which sometimes act as confounders in the analysis implying counter-intuitive spatial processes. Vastly different covariance structures can be observed for different values of the autoregressive parameter. This is worrying for area-based modelling and is similar in nature to that found in Openshaw and Taylor's (1979) analysis of voting data in Iowa US, where different arrangements of the spatial units yielded correlations from -1 to +1 for the same pre-aggregated data. For point-based data, geostatistical regressions are commonly used, where spatial dependencies are modelled directly via the variogram, a function specifying a deterministic relationship between point pairs and their correlation. Recent advances have also seen geostatistical models applied in health studies where area-based data is common (Goovaerts 2009). These models are worthy in that they do not suffer from the pitfalls described by Wall (2004).

Spatial information can also be accounted for by expressing the regression's coefficients as functions of the spatial coordinates (Casetti 1982; Gorr and Oligschaeger 1994), permitting a model of relationship nonstationarity. A major advance in this area was advent of the GWR model (Brunsdon et al. 1996). Here spatial structures in the data are incorporated into the model through a continuous distance-decay weighting scheme. The spatial extent of the scheme is controlled through the bandwidth parameter, which is optimised to give the best model fit to the data. The resultant localised regression coefficients are mapped to display their spatial variation.  Brunsdon et al. (1998) and  Harris et al. (2015) provide some simple inferential mechanisms to determine whether the coefficients exhibit significant spatial variation; and other notable advances in the GWR method can be found in Nakaya et al. (2005), Wheeler (2007), Huang et al. (2010), Harris et al. (2011), Brunsdon et al. (2012) and Silva and Fotheringham (2015). However, a fully-coherent inferential structure of GWR remains to be developed. It may be that, given GWR's kernel smoothing roots (e.g. see Cleveland 1979), the technique is best reserved for use as an exploratory and highly visual tool only, to act as a valuable precursor to a more sophisticated spatially-varying coefficient model, such as those proposed by Gelfand et al. (2003) or Assunção (2003).

The over-riding issue is that dependence and some form of nonstationarity are endemic in spatial data, and model forms should be made widely available which allow users to explore these structures both pre- and post-modelling. When spatial non-stationarity holds, the predictor and response variables is expected to change over space in an area.  Under spatial non-stationarity the relationship between the distribution of a certain species and the predictors shaping it may change over space. Thus any modelling activity  should also incorporate mechanisms to handle these characteristics of spatial data. This is an important challenge and this paper focuses on the modelling of spatial nonstationarity via the GWR model in order to demonstrate how to address process heterogeneity in relationships. The GWR model itself is formally described in Section 3, both in a basic form and an adapted form to deal with potential problems of localised collinearity in the predictor variables (Brunsdon et al. 2012; Gollini et al. 2015).

@@

Data recorded between 1980 and 2010 describing the presence of *Acer campestre* was downloaded from GBIF using the `dismo` R package, and subsetted for the UK. The data contained 22,701 records whose spatial distribution are shown in Figure 1, with a median of `r summary(colSums(res.vec))[3]`  records per year, a 1st quartile of `r summary(colSums(res.vec))[2]`  and a `r summary(colSums(res.vec))[4]` quartile of 917 records. The data for all years were summed over Ordnance Survey 10km grids because of the uneven distribution in time and space. Potential alternative approaches including generating absence points, background data (Phillips et al. 2009) to charcaterise study area environments or pseudo-absences (e.g. VanDerWal et al., 2009), indicating where absences might occur. However, pseudo absence approaches require a number of assumptions and lack statistical methods for handling the overlap between presence and background points (Ward et al. 2009; Phillips and Elith, 2011), absence data may be biased and / or incomplete (Kery et al., 2010) and background data approaches generate the same measures irrespective of where the species is observed (Hijmans and Elith, 2015).

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. In this study rainfall, Potential evapotranspiration (PET) and data describing a wildness quality index were used to construct a series of models for predicting the density of *Acer campestre* occurrence. Data on rainfall were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) which provides monthly 1km estimated rainfall data for each year. The average Autumn (3 month) rainfall was calculated for each 1km. Mean annual PET data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) which provides global data at 0.0083 degrees, approximately 1km. Finally, a wilderness quality dataset was included in the model. This was to explore the degree to the presence of *Acer campestre* may be related to anthropogenic disturbance – anecdotally this species is frequently used as an ornamental tree and found in field margins and hedges. The Wilderness Quality Index (WQI) data were generated for the whole of Europe at 1km resolution as described in Kuiters et al (2013). WQI can be considered as measure of non-anthropogenic activity and is based on a weighted linear combination of attributes related to wild land, wilderness and wildness (Carver et al, 2012, Comber et al, 2010) including naturalness of vegetation patterns, remoteness from settlements and human influence, and remoteness from roads. Each of these datasets were spatially aggregated over the OS 10m grid cells to generate mean values as shown in Figure 2.

@@

A multi-stage analysis was applied to model *Acer campestre* distributions. First, exploratory analyses were undertaken using a standard OLS regression to model distributions as an initial step. This identified significant predictor variables, under the assumption that relationships between predictor variables (rainfall, PET and wilderness) and species distributions are stationary (i.e. global). Then a GWR analysis was applied to examine the spatial variation in the relationships between the predictor variables and *Acer campestre* distributions (i.e. to test for the presence of local, non-stationary relationships). In overview, GW approaches use a moving window or kernel that passes through the study area. At each location being considered, data under the window are used to make a local calculation of some kind, such as a regression. The data are weighted by their distance to the kernel centre and in this way GW approaches construct a series of models at discrete locations in the study area. This is in contrast to global models, that consider all of the data (usually) in a single analysis of all data in the study area.

Next, the presence of global and local collinearity amongst the predictor variables was tested. This is a critical step in any regression, but for a GWR analysis it is usually overlooked. Collinearity occurs when variables exhibit linear or near linear relationships. Strong collinearity will affect model reliability and precision, generate unstable parameter estimates, inflated standard errors and inferential biases (Dormann et al 2013), and there may be problems in separating variable effects (Meloun et al. 2002). In a GWR analyses, collinearity may occur locally, with the construction of localised regressions, even when it is not observed globally (Wheeler and Tiefelsdorf, 2005; Wheeler 2007, 2009, 2013; Brunsdon et al 2012). A number of approaches exist to address collinearity in regression modelling, such as partial least squares regression, principal component analysis regression and ridge regression (Hoerl 1962; Hoerl and Kennard 1970). Ridge regression is a penalised model where extensions, such as the lasso and the elastic net also provide predictor variable sub-set selection (e.g. Zou and Hastie 2005). All such models could be adapted to a localised form and Wheeler (2007; 2009) has proposed both ridge and lasso versions of GWR to address any detrimental local collinearity effects. A related, locally-compensated ridge GWR model (LCR-GWR) is detailed in Brunsdon et al (2012), Lu et al. (2014) and Gollini et al (2015). It has advantages over the ridge GWR model of Wheeler (2007) in that a ridge term is applied locally and not globally. These studies also describe associated local collinearity diagnostics for GWR, such as the use of local correlations amongst pairs of predictors, local Variance Inflation Factors (VIFs) for each predictor, local variance decomposition proportions (VDPs) and the local condition numbers (CNs). The key point about LCR-GWR, is that a local ridge term is only applied where it is needed – when the local CN is above a pre-specified value in this case 30, which is a standard heuristic.

This study undertook a GWR analysis, with a locally-compensated ridge term if necessary, over a 200m grid of points covering the study area, with the aim of examining the spatial distribution of coefficient estimates predicting *Acer campestre* distributions and their spatial variation. Euclidean distances were used to weight data points under the kernel. These distances better reflect the spatial processes and relationships in environmental systems than network distance (Comber et al, 2008). For the kernel, an adaptive bi-square weighting function was applied, although a number of kernel functions can be specified for GW models as discussed in Gollini et al (2015). This generates higher weights at locations very near to the kernel centre relative to those towards the edge. For each data point ($P_j$) under the kernel (with a given bandwidth), a weight $w_{i,j}$ is calculated based on its distance to the centre of the kernel ($K_{i}$) as follows:

@@

An optimum kernel bandwidth for GWR can be found by minimising a model fit diagnostic. Options include a leave-one-out cross-validation (CV) score (Bowman 1984; Brunsdon et al. 1996). This optimises model prediction accuracy and the Akaike Information Criterion (AIC) (Akaike 1973; Fotheringham et al. 2002) optimises model parsimony by trading off prediction accuracy and complexity. In this case, the CV approach was applied to specify all GWR models, all using the bi-square weighting kernel and distances between locations.

@@

where $y_i$ is the response variable at location $i$, $x_{ik}$ is the value of the $k^{th}$ predictor variable at location $i$, $m$ is the number of predictor variables, $\beta_{i0}$ is the intercept term at location $i$, $\beta_{ik}$ is the local regression coefficient for the $k^{th}$ predictor variable at location $i$ and $\epsilon_i$ is the random error at location $i$. The result of the weighting means that data nearer to the kernel centre make a greater contribution to the estimation of local regression coefficients at each local regression calibration point $i$.

@@

A standard OLS regression models was undertaken and the resultant coefficient estimates and significance values are shown in Table 1 below. PET and Wilderness were found to be significant predictors of *Acer campestre* distributions at the 5% level (i.e. with a less than 95% chance of occurring randomly). Interestingly, in contrast to the findings of Coudun et al (2006), mean Autumn rainfall was not found to be significantly associated with the *Acer campestre* distributions.

@@

The underlying theoretical framework provided by GWR tests for non-stationarity in processes and relationships between factors. A standard GWR analysis was undertaken and in this case an optimal adaptive bandwidth of `r bw2` data points was determined using a cross-validation procedure.

The local coefficient estimates from this GWR model are shown in Table 2 and the significant variables, PET and WQI are mapped in Figures 3 and 4, respectively. They indicate considerable variation around the median in the degree to which increases in the predictor variables are associated with *Acer campestre* distributions. For example, considering the inter-quartile ranges shows that, in some places:

@@

Globally, the Variance Inflation Factors (VIFs) are all less than 10, although 2 of the Variance Decomposition Proportions (VDPs) are greater than 0.5 and the Condition Number (CN) is greater than 30, using standard heuristics from Belsley et al (1980) and O’Brien (2007). Together these suggest the presence of variable collinearity when evaluated globally, considering all data points together. Applying GWR collinearity diagnostics to the GWR model above generates local VIFs, local VDPs and local CNs at the same scale (i.e. using the same adaptive bandwidth of 21 data points). The results in Table 3 indicate a high degree of *local* collinearity in the GWR model. These values suggest that the application of a LCR-GWR is warranted. The local collinearity measures are mapped in Figure 5.

@@

Having tested for and identified local collinearity, a LCR-GWR was specified. This applies a GW regression but with a locally-compensated ridge term and fits local ridge regressions with their own ridge parameters (i.e., the ridge parameter varies across space), but only does this at locations where the local CN is above a user-specified threshold. In this case the CN threshold was specified as 30. An optimal adaptive bandwidth of `r lcr.bw` data points was again determined using a cross-validation procedure. Figures 6 and 7 show the spatial distribution of the original GWR coefficients, those determined using a LCR-GWR and a map of the differences between the two, for PET and for Wilderness Quality Index. In both cases there are large and potentially important differences between the coefficient estimates from the GWR and those from the LCR-GWR.

@@

# 5. Discussion

In this paper a series of analyses were undertaken to demonstrate the application and value of explicitly spatial analyses, focusing on GWR, and the need to reconsider common assumptions in a-spatial data analyses. Local statistical models were developed in order to test for spatial non-stationarity, in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere. Additionally, the paper highlights the importance of considering and testing for local collinearity especially in spatial non-stationarity models such as GWR, even where none is found to exist globally. In this analysis very strong evidence for local collinearity was found when the data were tested using local collinearity diagnostics. In all applications of GWR published in the literature, this critical step has been missed. Where local collinearity is found, locally-compensated models such as LCR-GWR can be applied and Brunsdon et al. (2012) propose a variety of these. A LCR-GWR only fits local ridge regressions at locations where the local Condition Number is above a user-specified threshold. Gollini et al (2015) review LCR-GWR with respect to alternative approaches for handling collinearity, but whatever adapted GWR model is applied, they all have the potential provide more accurate local coefficient estimates in the presence of collinearity than that found with a standard GWR model (Brunsdon et al. 2012).

@@

Akaike H (1973). Information Theory and an Extension of the Maximum Likelihood Principle. In BN Petrov, F Csaki (eds.), *2nd Symposium on Information Theory*, pp. 267–281. Akademiai Kiado, Budapest.

Anselin L (1988). *Spatial Econometrics: Methods and Models*, Dordrecht: Kluwer Academic 
Assunção R, 2003, Space varying coefficient models for small area data, *Environmetrics*, 14: 453-473

@@

Carver S, Comber A, McMorran R and Nutter S. (2012). A GIS model for mapping spatial patterns and distribution of wild land in Scotland. *Landscape and Urban Planning*, 104 (2012) 395–409.

@@

Comber AJ, Carver S, Fritz S, McMorran R, Washtell J and Fisher P. (2010). Different methods, different wilds: evaluating alternative mappings of wildness using Fuzzy MCE and Dempster Shafer MCE. *Computers, Environment and Urban Systems*, 34: 142-152

@@

Fotheringham, A. S., Brunsdon, C., & Charlton, M. (2002). Geographically weighted regression: the analysis of spatially varying relationships. John Wiley & Sons.

>>>>

---
title: 'Spatial considerations in the analysis of biological conservation data: space is special'
author: "Alexis Comber^1^, Steve Carver^1^, Carol Ximena Garzon-Lopez^2^, Paul Harris^3^, Duccio Rocchini^4^"
output: pdf_document
---

@@

# Abstract
Geographically Weighted Regression (GWR) as first proposed by Brunsdon et al (1996) is a commonly used approach in many local geographical and biogeographical analyses. One of the of criticisms of the use of GWR is that local models may be subject to local collinearity between variables, even when none is observed globally. As yet none of the published research describing applications of GWR has addressed this issue. This paper does so. It applies GWR to the model suggested by Coudun et al (2006) to predict the presence of *Acer campestre* and to account for any spatial non-stationarity, as observed in many statistical patterns and relationships. The analysis tests for presence of local collinearity amongst predictor variables and applies a locally compensated GWR model where it is needed. The results describe the spatial variation of coefficient estimates predicting *Acer campestre* and compensates for any local collinearity amongst variables using a local ridge term. The paper discusses the need for *geography goggles* in  analyses of spatial and in so doing it encourages explicitly spatial ways of thinking. Such considerations are increasingly relevant as more and more data have a spatial component in the form of location, for instance from GPS, and all data are collected some*where*. These suggest the need for more informed approaches for analysing space and location than provided by standard statistical models.

@@

This paper emphasises the importance of explicitly considering the spatial properties of ecological data and argues that location cannot be treated as just another variable because of the prevalence of spatial autocorrelation or spatial non-stationarity. To demonstrate the value of spatially explicit methods, this paper spatially extends the model suggested by Coudun et al (2006) predicting the presence of *Acer campestre* using a Geographically Weighted Regression (GWR) framework (Brunsdon et al, 1996). The extension accounts for a particular form of spatial dependence when the properties of nearby features are found to be correlated, contradicting the underlying assumptions of independence and stationarity made in many statistical analyses. The result is spatial autocorrelation or spatial non-stationarity.

Spatially explicit approaches such as GWR  quantify *local* patterns in relationships and processes rather than global, 'whole map' statistics. Typically, local models are constructed from subsets of the data, for example by selecting data under a moving window or kernel, and GWR has been used in a number of ecological studies. Examples include Roll et al (2015) who examined the spatial variation tree height as a predictor of species richness for different taxa, Wang et al (2016) who used to GWR to develop local models of net primary production and Keith et al (2013) who sought to predict local species richness using a GWR analysis of environmental variables. However, local methods may be subject to collinearity, even when it is not observed globally (Brunsdon et al 2012). Collinearity affects model reliability and can result in inferential bias through unstable parameter estimates, inflated standard errors and difficulties in separating variable effects (Dormann et al 2013; Meloun et al. 2002). Despite many papers describing applications of GWR, there are no instances where the potential effects of local collinearity have been considered. As well as extending the analysis of Coudun et al (2006) using GWR, this paper explicitly tests for and compensates for locally collinearity where found, with dramatic differences between the coefficient estimates and the mapped GWR outputs without and without local compensations.

@@

Understanding species distribution and their spatial dependencies is a key topic in ecological research (Pearman et al., 2008; Rocchini et al., 2015) and biodiversity monitoring (Honrado et al., 2016). Reliable descriptions of distributions are fundamental for conservation and research purposes (Dormann, 2007) where a lack of spatial sensitivity hampers the ability to provide coherent biodiversity management strategies (Guisan et al., 2014, Porfirio et al., 2014). However, when modelling species distributions, a number of statistical problems must be overcome including the uncertainty related to sampling (Rocchini et al., 2011), correlation among predictors (Dormann et al., 2012), the spatial variability of predictors (Rocchini et al., 2016) and their spatio-temporal dependencies (Zuur and Ieno, 2016). Spatial non-stationarity and scale dependence have been found to have the greatest impact on observed ecological processes and patterns (Foody, 2004).

The importance of spatial considerations has a long history in a number of disciplines: Fischer (1935) in crop science; Kolmogorov (1941), Gandin (1965) in meteorology; Krige (1951) and  Matheron (1963) in mining; Matérn (1960) in forestry; theoretical developments by Moran (1950) and Yaglom (1955); Berry and Marble (1968) and Chorley and Haggett (1967) in geography; and Legendre and Legendre (1991) in ecology. **The** major issue with modelling spatial data is the lack of independence of the spatial objects and the assumption of stationarity in the processes being modelled. These are well recognised in socio-economic domains, but are also relevant within bio-geographical, environmental and ecological ones and are reflected in Tobler's dictum (the First Law of Geography) that: '*everything is related to everything else, but near things are more related than distant things*' (Tobler 1970). However, despite numerous methodological advances in addressing such spatial modelling problems, whose lineage from the late 1970s can be loosely traced through spatial statistics texts from Journel and Huigbrechts (1978) to Cressie (1993), to Chilès and Delfiner (1999) and to Cressie and Wilke (2011), policy related research is still routinely informed by non-spatial modelling practices which are heroic at best, and ill-advised at worst.

Focusing on regression models, the main drawback in assuming observations are independent is that any spatial dependencies turn up in the residuals, as they are not explicitly added to the model. For area-based data, ways to account for this include Besag's (1974) conditionally autoregressive model or Anselin's (1988) spatially autoregressive model. For point-based data, geostatistical regressions are commonly used, where spatial dependencies are modelled directly via the variogram with a number of recent advances (Goovaerts 2009). Spatial information can also be accounted for by expressing the regression's coefficients as functions of the spatial coordinates (Casetti 1982; Gorr and Oligschaeger 1994), permitting a model of relationship nonstationarity. GWR (Brunsdon et al. 1996) was a major advance in this area. It uses a continuous distance-decay weighting scheme to accommodate the spatial structures in the data and the resultant localised regression coefficients can be mapped. Brunsdon et al. (1998) and  Harris et al. (2015) provide some simple inferential mechanisms to determine whether the coefficients exhibit significant spatial variation. Some of the advances in the GWR method can be found in Nakaya et al. (2005), Wheeler (2007), Huang et al. (2010), Harris et al. (2011), Brunsdon et al. (2012) and Silva and Fotheringham (2015).

In summary, a key issues in ecological modelling is that dependence and some form of nonstationarity are endemic in spatial data. When spatial non-stationarity occurs the predictor and response variables are expected to change over space and the relationship between the distribution of a certain species and predictor variables may change over space. Mechanisms to handle these characteristics of spatial data in modelling is an important challenge. This paper accommodates spatial nonstationarity using the GWR model in order to demonstrate how to address process heterogeneity in relationships. The GWR model itself is formally described below, in both basic and an adapted forms to deal with potential problems of localised collinearity in the predictor variables (Brunsdon et al. 2012; Gollini et al. 2015).

@@

Data of *Acer campestre* presence (1980-2010) for the UK was downloaded from GBIF using the `dismo` R package. It contained 22,701 records whose spatial distribution is shown in Figure 1. The data for all years were summed over Ordnance Survey 10km grids rather than generating absence points and background data (Phillips et al. 2009; VanDerWal et al., 2009) which require a number of assumptions (Ward et al. 2009; Phillips and Elith, 2011) and may be biased (Kery et al., 2010; Hijmans and Elith, 2015).

@@

The study by Coudun et al (2006) found the presence of *Acer campestre* to be significantly related to Autumn rainfall and actual Thornthwaite evapotranspiration. Rainfall data were downloaded from the NERC Environmental Information Data Centre (Tanguy et al, 2015) and the average Autumn (3 month) rainfall was determined for each 1km. Mean annual potential evapotranspiration data were downloaded from the CGIAR Consortium for Spatial Information (Trabucco and Zomer, 2009) at 0.0083 degrees (approximately 1km). Finally, a wilderness quality dataset was included in the model to explore how the presence of *Acer campestre* was related to anthropogenic disturbance as anecdotally it is frequently planted as an ornamental tree and found in hedges. The European Wilderness Quality Index (WQI) (Kuiters et al 2013) provided a measure of non-anthropogenic activity. Each of these datasets were spatially aggregated over the OS 10m grid (Figure 2).

@@

A multi-stage approach was used to model *Acer campestre* distributions. First, a standard OLS regression to model distributions was used to identify significant predictor variables, under the assumption that relationships between predictor variables (rainfall, PET and wilderness) and species distributions are stationary (i.e. global). Second, a GWR analysis was applied to examine the spatial variation in these relationships, testing for the presence of local, non-stationarity. In overview, GW approaches use a moving window or kernel that passes through the study area. At each location being considered, data under the window are used to make a local calculation of some kind, such as a regression. The data are weighted by their distance to the kernel centre and in this way GW approaches construct a series of models at discrete locations in the study area. Third, the presence of global and local collinearity amongst the predictor variables was tested for. This is a critical step in any regression, but for a GWR analysis it is usually overlooked. Collinearity occurs when variables exhibit linear or near linear relationships. In a GWR analyses, collinearity may occur locally, with the construction of localised regressions, even when it is not observed globally (Wheeler and Tiefelsdorf, 2005; Wheeler 2007, 2009, 2013; Brunsdon et al 2012). A number of approaches exist to address collinearity in regression modelling, such as partial least squares regression, principal component analysis regression and ridge regression (Hoerl 1962; Hoerl and Kennard 1970). Ridge regression is a penalised model where extensions, such as the lasso and the elastic net also provide predictor variable sub-set selection (e.g. Zou and Hastie 2005). A locally-compensated ridge GWR model (LCR-GWR) is detailed in Brunsdon et al (2012), Lu et al. (2014) and Gollini et al (2015) as are associated local collinearity diagnostics for GWR. These include local correlations amongst pairs of predictors, local Variance Inflation Factors (VIFs) for each predictor, local variance decomposition proportions (VDPs) and  local condition numbers (CNs). The LCR-GWR only applies a local ridge term where it is needed: in this case when the local CN is above a pre-specified threshold of 30, which is a standard heuristic. So in summary, this study applied an OLS regression, compared this with a GWR analysis and then applied a locally-compensated ridge term if necessary. The local GWR analyses were undertaken at discrete locations in a hexagonal grid of 3028 points covering the study area.

For the geographically weighted kernels, an adaptive bi-square weighting function was applied. This generates higher weights at locations very near to the kernel centre relative to those towards the edge. For each data point ($P_j$) under the kernel with a given bandwidth, a weight $w_{i,j}$ is calculated based on its distance to the centre of the kernel ($K_{i}$) as follows:

@@

Optimum kernel bandwidths were found for all of the GWR models by minimising a model fit diagnostic using a leave-one-out cross-validation (CV) score (Bowman 1984; Brunsdon et al. 1996).

@@

where $y_i$ is the response variable at location $i$, $x_{ik}$ is the value of the $k^{th}$ predictor variable at location $i$, $m$ is the number of predictor variables, $\beta_{i0}$ is the intercept term at location $i$, $\beta_{ik}$ is the local regression coefficient for the $k^{th}$ predictor variable at location $i$ and $\epsilon_i$ is the random error at location $i$. The weighting results in data nearer to the kernel centre making a greater contribution to the estimation of local regression coefficients at each local regression calibration point $i$.

@@

The coefficient estimates from a standard OLS regression are shown in Table 1 below. PET and Wilderness were found to be significant predictors of *Acer campestre* distribution at the 5% level. Interestingly, in contrast to the findings of Coudun et al (2006), mean Autumn rainfall was not found to be significantly associated with the *Acer campestre* distributions.

@@

Next a standard GWR analysis was applied and an optimal adaptive bandwidth of `r bw2` data points was determined using a cross-validation procedure. The *local* coefficient estimates from this GWR model are shown in Table 2 and the significant variables, PET and WQI are mapped in Figures 3 and 4, respectively. They indicate considerable variation around the median in the degree to which increases in the predictor variables are associated with *Acer campestre* distributions. For example, considering the inter-quartile ranges shows that, in some places:

@@

Globally, the Variance Inflation Factors (VIFs) are all less than 10, although 2 of the Variance Decomposition Proportions (VDPs) are greater than 0.5 and the Condition Number (CN) is greater than 30, using standard heuristics from Belsley et al (1980) and O’Brien (2007). These suggest the presence of global variable collinearity. Applying GWR collinearity diagnostics to the GWR model above generates local VIFs, local VDPs and local CNs at the same scale (i.e. using the same adaptive bandwidth of `r bw2` data points). The results in Table 3 indicate a high degree of *local* collinearity in the GWR model. These values suggest that the application of a LCR-GWR is warranted. The local collinearity measures are mapped in Figure 5.

@@

Having tested for and identified local collinearity, a LCR-GWR was specified. This applies a GW regression but with a locally-compensated ridge term and fits local ridge regressions with their own ridge parameters (i.e., the ridge parameter varies across space), but only does this at locations where the local CN is above a user-specified threshold. In this case the CN threshold was specified as 30. An optimal adaptive bandwidth of `r lcr.bw` data points was again determined using a cross-validation procedure. Figures 6 and 7 show the spatial distribution of the original GWR coefficients, those determined using a LCR-GWR and a map of the differences between the two, for PET and for WQI. In both cases there are large and potentially important differences between the coefficient estimates from the GWR and those from the LCR-GWR.

@@

# 5. Discussion and Conclusions

In this paper a series of analyses were undertaken to demonstrate the application and value of explicitly spatial analyses, focusing on GWR, and the need to reconsider common assumptions in a-spatial data analyses. Local statistical models were applied to test for spatial non-stationarity, in contrast to standard, a-spatial, statistical approaches that assume the relationships between factors to be the same everywhere. Then a locally-compensated ridge GWR was used to handle any local collinearity in the GWR models, with large differences in outcome. The results highlights the importance of considering and testing for local collinearity especially in spatial non-stationarity models such as GWR, even where none is found to exist globally. This critical step has been missed in all GWR applications published to date and collinearity-aware methods were found to provide more accurate local coefficient estimates in the presence of collinearity. Gollini et al (2015) provide a review of approaches for handling local collinearity.

@@

Anselin L (1988). *Spatial Econometrics: Methods and Models*, Dordrecht: Kluwer Academic

@@

@@

@@
